---
title: "Cruise Line Customer Lifetime Value: BG / BB Model"
author: "Brian Bates, Zach Dorflinger, Kristina Paterson, Chris Stocks"
date: "2/6/2020"
output: html_document
---

```{r, include=F, warning=F}
library(dplyr)
library(ggplot2)
library(xtable)
library(tidyr)
library(knitr)
library(kableExtra)
library(BTYD)
library(scales)

options(digits = 3)
options(xtable.comment = F)
```

```{r}
# Data Import / Cleaning
# Calibration
raw_calib <- read.delim("./cruises_calib.txt", header = T, sep = "")

data_calib <- as.data.frame(t(spread(raw_calib,key=ID,value=1,fill=0))) %>% 
    setNames(c("ID","2010","2011","2012","2013","2014")) %>% 
    slice(-1) %>% 
    mutate_at(vars(-ID), function(x) {ifelse(x>0,1,0)}) %>% 
    mutate(
        X = rowSums(.[-1]),
        tx = ifelse(`2014` > 0, 5,
                    ifelse(`2013` > 0, 4,
                           ifelse(`2012` > 0, 3,
                                  ifelse(`2011` > 0, 2,
                                         ifelse(`2010` > 0, 1, 0)))))
    )

summary_calib <- data_calib %>% 
    group_by(X) %>% 
    summarise(Count = n())

# Holdout
raw_hold <- read.delim("./cruises_hold.txt", header = T, sep = "")

data_hold <- t(spread(raw_hold,key=ID,value=1,fill=0))
data_hold <- data.frame(names = row.names(data_hold), data_hold) %>% 
    setNames(c("ID","2015","2016","2017","2018")) %>% 
    slice(-1) %>% 
    mutate_at(vars(-ID), function(x) {ifelse(x>0,1,0)}) %>% 
    mutate(
        X = rowSums(.[-1]),
        ID = as.numeric(as.character(ID))
    )

# Combined
data_full <- left_join(x = select(data_calib, -c(X, tx)),
                       y = select(data_hold, -c(X)),
                       by = "ID") %>% 
    replace(is.na(.), 0) %>% 
    mutate(
        X = rowSums(.[-1]),
        tx = ifelse(`2018` > 0, 9,
                ifelse(`2017` > 0, 8,
                    ifelse(`2016` > 0, 7,
                        ifelse(`2015` > 0, 6,
                            ifelse(`2014` > 0, 5,
                                ifelse(`2013` > 0, 4,
                                    ifelse(`2012` > 0, 3,
                                        ifelse(`2011` > 0, 2,
                                            ifelse(`2010` > 0, 1, 0)))))))))
    )

summary_holdout <- data_full %>% 
    select(ID, `2015`:`2018`) %>% 
    mutate(X = rowSums(.[-1])) %>% 
    group_by(X) %>% 
    summarise(Count = n())

summary_full <- data_full %>% 
    group_by(X) %>% 
    summarise(Count = n())

# Automated way to calculate tx. Works, but is MUCH slower than if statements.
# numeric_years <- as.numeric(colnames(select(data_full, -c(ID, X, tx))))
# for (i in 1:nrow(data_full)) {
#     data_full[i, "tx"] = max(numeric_years * select(data_full, -c(ID, X, tx)))
# }
```

### 1. Estimating the BB and BG-BB models 

#### Beta-Binomial

```{r}
# Model Estimation
log_bb <- function(x,m,a,b) {
  lchoose(m,x) + lbeta(a+x,b+m-x) - lbeta(a,b)
}

LL_BB <- function(pars, x, m) {
  a <- exp(pars[1]) 
  b <- exp(pars[2])
  z <- log_bb(x,m,a,b)
  return(-sum(z))
}

start <- log(c(1,1))
opt <- optim(start, LL_BB, x = data_calib$X, m=5)
```

```{r}
# Parameter Estimates
a <- exp(opt$par[1])
b <- exp(opt$par[2])

LL <- opt$value

avg_theta <- a / (a + b)
var_theta <- (a*b) / (((a + b)^2) *(a + b + 1))
```

```{r}
# Probability density plot
ggplot(tibble(theta=c(.001,.999)), aes(x=theta)) %>% +
  stat_function(fun=dbeta, 
                args=list(shape1=a,shape2=b), 
                geom="line", color = "blue") %>% + 
  scale_x_continuous(expression("Probability of Cruising") , breaks= seq(0,1,.1)) %>% + 
  scale_y_continuous("Proportion of Customers") %>% +
  geom_vline(xintercept=avg_theta, linetype=5) %>% + 
  theme(axis.text.y=element_blank())
```

The parameter estimates for the BB model are the following: 

* $\alpha$: `r round(a, 3)`
* $\beta$: `r round(b, 3)`
* Log-likelihood: `r comma_format()(round(LL, 1))`
* Avg cruise probability ($\theta$): `r round(avg_theta*100, 1)`%
* Variance of cruise probability: `r round(var_theta*100, 1)`%

According to this model, the typical customer in the data has a 9% likelihood
of taking a cruise in any given year. There is variation in this likelihood, 
however, and most customers have very low likelihoods of cruising in a given
year. Very few customers are more likely than not to cruise in any specific 
year.

#### Beta-Geometric / Beta-Binomial

```{r}
# Model Estimation
get_bgbb_LL <- function(pars, x, tx, m) {
  p <- exp(pars)
  bg_LL <- bgbb.LL(p, x, tx, m)
  return(-sum(bg_LL))
}
st <- rep(0,4)

bgbb_opt <- optim(st, get_bgbb_LL,
                  x=data_calib$X,
                  tx=data_calib$tx,
                  m=5,
                  method = "BFGS")
```

```{r}
# Parameter Estimates
bgbb_pars <- exp(bgbb_opt[['par']])
bgbb_LL <- bgbb_opt[['value']]

bgbb_a <- bgbb_pars[1]
bgbb_b <- bgbb_pars[2]

avg_bgbb_p <- bgbb_a / (bgbb_a + bgbb_b)
var_bgbb_p <- (bgbb_a*bgbb_b) / (((bgbb_a + bgbb_b)^2) *(bgbb_a + bgbb_b + 1))

bgbb_c <- bgbb_pars[3]
bgbb_d <- bgbb_pars[4]

avg_bgbb_theta <- bgbb_c / (bgbb_c + bgbb_d)
var_bgbb_theta <- (bgbb_c*bgbb_d) / (((bgbb_c + bgbb_d)^2) *(bgbb_c + bgbb_d + 1))
```

```{r}
# Cruise probability density
ggplot(tibble(theta=c(.01,.99)), aes(x=theta)) %>% +
  stat_function(fun=dbeta, 
                args=list(shape1=bgbb_a,shape2=bgbb_b), 
                geom="line", color = "blue") %>% + 
  scale_x_continuous(expression("Probability of Cruising") , breaks=seq(0,1,.1)) %>% + 
  scale_y_continuous("Proportion of Customers") %>% +
  geom_vline(xintercept=avg_bgbb_p, linetype=3) %>% + 
  theme(axis.text.y=element_blank()) +
  ggtitle("Variation in Cruise Probability")

# Death probability density
ggplot(tibble(theta=c(.01,.99)), aes(x=theta)) %>% +
  stat_function(fun=dbeta, 
                args=list(shape1=bgbb_c,shape2=bgbb_d), 
                geom="line", color = "blue") %>% + 
  scale_x_continuous(expression("Probability of Dropping") , breaks=seq(0,1,.1)) %>% + 
  scale_y_continuous("Proportion of Customers") %>% +
  geom_vline(xintercept=avg_bgbb_theta, linetype=3) %>% + 
  theme(axis.text.y=element_blank()) +
  ggtitle("Variation in Death Probability")

```

The parameter estimates for the BB/BG model are the following: 

* $\alpha$: `r round(bgbb_a, 3)`
* $\beta$: `r round(bgbb_b, 3)`
* c: `r round(bgbb_c, 3)`
* d: `r round(bgbb_d, 3)`
* Log-likelihood: `r comma_format()(round(bgbb_LL, 1))`
* Avg cruise probability: `r round(avg_bgbb_p*100, 1)`%
* Variance of cruise probability: `r round(var_bgbb_p*100, 1)`%
* Avg death probability: `r round(avg_bgbb_theta*100, 1)`%
* Variance of death probability: `r round(var_bgbb_theta*100, 1)`%

According to this model, the typical consumer has a 12% likelihood of taking
a cruise in any given year, conditional on the fact that they have not "died" or
decided not to take any cruises ever again. Most customers who are still alive
have a probability of cruising in the neighborhood of 5-30%. The likelihood of 
"death" for the average coustomer is 12.3% in any given year, however there
are a decent number of customers with mugh higher likelihoods. 

### 2. Model Performance In-sample

```{r}
# BB 
pred_in_sample <- data.frame("X" = 0:5, "bb_prob" = NA)
for (i in 0:5) {
  pred_in_sample$bb_prob[i+1] = exp(log_bb(i,5, a, b))
}
pred_in_sample$bb_estimate <- pred_in_sample$bb_prob * nrow(data_calib)

# BGBB
pred_in_sample$bgbb_prob <- bgbb.pmf(bgbb_pars, 5, 0:5)
pred_in_sample$bgbb_estimate <- pred_in_sample$bgbb_prob * nrow(data_calib)

# Actual
pred_in_sample <- merge(pred_in_sample, summary_calib, by="X")

# Formatting
table_in_sample <- pred_in_sample %>% 
    select(c(X, Count, bb_estimate, bgbb_estimate)) %>% 
    rename(
        "Num Cruises" = X,
        "Actual" = Count,
        "BB Estimate" = bb_estimate,
        "BGBB Estimate" = bgbb_estimate
    ) %>% 
    mutate_all(round)
```

```{r}
# Table
kable(table_in_sample) %>% 
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

```{r}
# Plot
graph_data <- gather(table_in_sample, Model, value, -`Num Cruises`)

ggplot(graph_data, aes(x = `Num Cruises`, y = value, fill = Model)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_x_continuous(breaks = seq(0, 5, 1)) +
    xlab('Total Transactions: 2010-2014') + 
    ylab('Number of Customers') + 
    ggtitle('Actual Customers vs Estimated: Calibration Period')
```

Both models appear to fit the data quite well during the calibration period
(2010-2014). There are slight differences between the model predictions 
and actual results, but overall, model performance is solid. This makes sense,
as the models are trained using this data and there are a small number of years
in the sample, so creating a model that matches the small amount of data nearly 
perfectly is not a challenge.

### 3. Model Performance in Holdout Period

```{r}
# BB 
pred_out_sample <- data.frame("X" = 0:4, "bb_prob" = NA)
for (i in 0:4) {
  pred_out_sample$bb_prob[i+1] = exp(log_bb(i,4, a, b))
}
pred_out_sample$bb_estimate <- pred_out_sample$bb_prob * nrow(data_calib)

# BGBB
pred_out_sample$bgbb_prob <- bgbb.pmf.General(bgbb_pars,5,4,0:4)
pred_out_sample$bgbb_estimate <- pred_out_sample$bgbb_prob * nrow(data_calib)

# Actual
pred_out_sample <- merge(pred_out_sample, summary_holdout, by="X")

# Formatting
table_out_sample <- pred_out_sample %>% 
    select(c(X, Count, bb_estimate, bgbb_estimate)) %>% 
    rename(
        "Num Cruises" = X,
        "Actual" = Count,
        "BB Estimate" = bb_estimate,
        "BGBB Estimate" = bgbb_estimate
    ) %>% 
    mutate_all(round)
```

```{r}
# Table
kable(table_out_sample) %>% 
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

```{r}
# Plot
graph_data <- gather(table_out_sample, Model, value, -`Num Cruises`)

ggplot(graph_data, aes(x = `Num Cruises`, y = value, fill = Model)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_x_continuous(breaks = seq(0, 5, 1)) +
    xlab('Total Transactions: 2015-2018') + 
    ylab('Number of Customers') + 
    ggtitle('Actual Customers vs Estimated: Holdout Period')
```

The BG/BB model predictions align much closer with the actual transaction levels
in 2015-2018 than the BB model. The BB model underestimates how many consumers 
took 0 cruises during the period, while the BG/BB did not. This is consistent
with the model assumptions, as the BG/BB model appears to be correctly diagnosing
customers that have "died" or dropped out of the sample while the BB model is not.

### 4. BG-BB Forecasting

#### A) Number of Surviving Customers

```{r}
probs_vector <- bgbb.PAlive(bgbb_pars, 
                            x = data_full$X, 
                            t.x = data_full$tx, 
                            n.cal = 9)
num_alive <- sum(probs_vector)
```

According to the estimated BG/BB model, we expect `r comma_format()(round(num_alive))`
customers to still be alive at the beginning of 2019.

#### B) Total Expected Residual Lifetime Value

```{r}
dert_vector <- bgbb.DERT(bgbb_pars, 
                         x = data_full$X, 
                         t.x = data_full$tx,
                         n.cal = 9,
                         d = 0.13)
total_value <- 241 * sum(dert_vector)
```

The total expected residual lifetime value of the 2009 cohort at the beginning
of 2019 is $`r comma_format()(round(total_value))`.

### 5. Potential New Customer Acquisition

#### A) Expected Cruises from 2020-2024

```{r}
forecast <- data.frame(
    "Year" = 2020:2024,
    "Est.Cruises" = round(bgbb.pmf(bgbb_pars, 5, 0:4) * 11993)
)

kable(forecast) %>% 
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

#### B) Max Willingness to Pay

```{r}
# DERT of a new customer with no prior info
blind_dert <- bgbb.DERT(bgbb_pars, 
                        x = 0,
                        t.x = 0,
                        n.cal = 0,
                        d = 0.13)

# Num expected transactions (DERT + guaranteed one) * customers * margin
value <- (blind_dert + 1) * 11993 * 241
```

The estimated value of the 11,993 new customers is $`r comma_format()(round(value))`.
The company should not spend more than this amount on the campaign.
